Can Symmetric Cryptography Be Liberated from the Von Neumann Style?

### 2024-10-12 paper.tex
Add refs

Anyone who has ever read a cryptography paper knows this for a fact:
cryptographers are at the top of the academic thought chain. In the
archetypal symmetric cryptography paper, there is a ``Section 2''
scene in which cryptographers are bravely waving their most potent
finite field. In a latter ``Section 4'' scene, the very same
cryptographers are seen extolling the virtues of some nifty
\textsc{Arm} Cortex assembly instruction, which takes center stage in
their hand-crafted, manually optimized implementation. No wonder we
jealously stash them away in ``zone à régime restrictive''! The rumor
goes that, behind these padlocked doors, they have developed a set of
domain-specific techniques to deliver very high-throughput software
implementation of symmetric ciphers. This has been slowly gnawing at
compiler writers, for whom job security~\citep{bernstein} has become a
major concern. Luckily for us, every now and then, someone
inadvertently let the genies out of their Klein bottle and we get to
read about the subtle art of ``bitslicing''~\citep{kwan:bitslicing}

@@

In the long run, our ambition is to offer cryptographers the
conceptual apparatus to better illuminate their ``Section 5'' scene.
Each and every symmetric cryptosystem should not have to exhibit a
bitsliced or fixsliced implementation: by just \emph{specifying} the
cryptosystem in the right framework (abstract yet suitably
operational), one should get the guarantee that the design can be
bitsliced and fixsliced. In fact, one would boldly claim that a
compiler could automatically apply these transformations and produce
optimized code, starting from an high-level description of the cipher.
This is already the case for bitslicing~\citep{usuba}, adding support
for fixslicing is next on the list.

@@

To avoid side-channel attack (for example, based on
timing~\citep{time-side-channel}), this pipeline can be more precisely
understood as a combinational circuit. We are forbidden from
performing control-flow operations on secret data, leaving only
statically-bounded loops. We are also forbidden from performing memory
access based on the data flow, leaving only register spilling.

@@

Following the NSA involvement in the design of the DES diffusion
layer~\citep{des-history}, cryptographers have been advocating for
``nothing-up-my-sleeve'' designs (which does not make magic tricks
impossible, just harder~\citep{bernstein:something-up-my-sleeves}).
For example, the diffusion layer of the AES standard is based on the
usual multiplicative inverse over the finite field induced by an
innocent-looking polynomial. To support such design, the state of
modern ciphers is thus equipped with some matricial structure
(oftentimes, manipulated as a 2D or 3D array of bits): the cipher is
not specified over a flat sequence of individual bits but over a
n-dimensional array of bits.

@@

Moreover, ciphers are now, by and large, designed to be run
efficiently on software. The era of hardware-based cryptography is
long gone, even for embedded platforms~\citep{lightweight-crypto}. To
achieve high-throughput, ciphers are typically designed so as to be
admit a bitsliced implementation. Bitslicing enables a form of
parallel execution \emph{within} machine words. We thus gain access to
a form of CPU-level ``scale out'' to larger machine words, in
particular through single-instruction multiple-data (SIMD)
instructions. Doubling the size of machine words yields nearly twice
the throughput, without any effort. Evidences tend to suggest that
cryptographers obtain bitsliced designs through sheer intellectual
might. It is in fact the sole purpose of the ``Section 5'' scene to
exhibit a bitsliced witness in all the gory details.

@@

We believe that there is also a broader take-away for an audience of
stubbornly French functional programmers. The present work surfs on
the wake of the Squiggol school of
programming~\citep{wg21,gibbons:naperian}. Because our
government-approved programming language does not yet support ad-hoc
polymorphism~\citep{remy:explicit-poly}, the French community may be
missing out on some interesting programming patterns. The present
article can thus be read as a case-study in ``functor-oriented
programming'', where the word ``functor'' is taken in his categorical
meaning (recalled in Section~\ref{sec:spec} and unrelated to the one
mentioned in the ML module system). In effect, we forbid ourselves
from programming over a concrete datastructure: instead, we rely on
(categorical) functors, peppered with some more structure, so as to 1.
unlock algebraic manipulation during compilation and 2. exploit
parametricity for reasoning.

@@

The present work is but the beginning of our research program. We
shall focus \emph{exclusively} on the semantics aspects, only briefly
touching upon implementation aspects (which remain the raison d'être
of the project!). Our experience designing the Usuba~\citep{usuba}
programmign language and implementing its compiler gives us some
confidence that a syntax can indeed be tailored to suit this
semantics.

We nonetheless could not resist the temptation of developing the
semantics in the Coq theorem prover, as executable programs. After
all, we are interested in combinational circuits: if there has ever
been a time where we do not have to worry about termination, this is
it! So we have implemented cryptographical primitive in pure Gallina.
This is nothing to worry about: after all, more ambitious colleagues
went from there down to deeply-embedded assembly
code~\citep{chlipala}. Besides the ability to prove the correctness
and, even more usefully, test our code, this also enables us to easily
prototype a fixslicing compiler, using Coq's \texttt{autorewrite}
tactics to emulate a rule-based simplification
engine~\citep{haskell:simplier} (part of an imaginary optimizing
compiler).

@@

This article grew out of the Bachelor's research project of the first
author and the Master thesis of the second author. The former worked
on the specification, bitslicing and fixslicing of the Skinny
cipher~\citep{skinny-spec, adomin:skinny}. The latter worked on the
specification, bitslicing and fixslicing of the Gift
cipher~\citep{gift-spec, admon:gift}. Both ciphers are provided in the
aaccompanying source code~\footnote{\url{https://github.com/pedagand/bitfix}}. For clarity, we shall
focus exclusively on Skinny as our running example here.

@@

Instead, we require \coqe!Rows.T! to offer a \coqe!map! operation: it
constructively witnesses the fact that a \coqe!Rows.T A! is
(functionally) related to a \coqe!Rows.T B!, as long as we can
(functionally) relate \coqe!A! and \coqe!B!. Put otherwise, we ask
that \coqe!Rows.T!, and similarly \coqe!Cols.T! and \coqe!Slice8.T!,
come equipped with their free theorems~\citep{wadler:free-theorem}.

@@

Much like $\lambda$ in functional programming, the exclusive-or
features prominently in the cryptographic liturgy. In particular, it
is instrumental to mix the derived keys during the ciphertext
computation. For genericity, we model this layer as an operation
defined over any Boolean algebra~\citep{boolean-alg}:
%
\coqfrom{Skinny/Base}{addroundkey}

@@

The type \coqe!bool! is an obvious instance of a Boolean algebra.
However, any commutative applicative~\citep{applicative-functor,
  commutative-appfunc} functor \coqe!F! applied to a Boolean algebra
also yields a Boolean algebra (dispatching the Boolean operations
pointwise to the underlying elements). For the sake of completeness,
we recall that an applicative functor offers the following operations:
%
\coqfrom{Algebra}{Applicative}

@@

The role of the diffusion layer is to divert individual bits across the
whole structure. Inevitably, this calls for a notion of indexing.
Naperian functors~\citep{naperian} identify a type \coqe!Ix! of
indices as the logarithm of an exponential type:
%
\coqfrom{Algebra}{Naperian}
%
through the fact that \coqe!lookup! and \coqe!init! form a bijection.
This is also known as a representable functor in mathematical
circles~\citep{mclane:working-categorist}

@@

The first step of diffusion process consists in performing a right
rotation by a specific amount over each individual column. However, we
do not have enough structure to identify a ``right'' or ``left''
direction over our containers. The standard tool of the trade to
achieve this effect is the notion of foldable
functor~\citep{foldable}:
%
\coqfrom{Algebra}{Foldable}

@@

\begin{figure}[htp]
  \centering
  \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \includegraphics[draft=false,height=2.8cm]{shiftrows}
    \caption{Shiftrows (from~\citet{skinny-spec})}
    \label{fig:shiftrows}
  \end{subfigure}
  \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \includegraphics[draft=false,height=2.8cm]{mixcolumns}
    \caption{Mixcolumns (from~\citet{skinny-spec})}
    \label{fig:mixcolumns}
  \end{subfigure}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[draft=false,height=5cm]{s8}
  \caption{Substitution box S8 (from~\citet{skinny-spec})}
  \label{fig:s8}
\end{figure}

@@

Given two Naperian functors \coqe!F! and \coqe!G!, there exists a
generic notion of reindexing~\citep{gibbons:apl-icative}:
%
\coqfrom{Algebra}{reindex}

@@

Studying the Gift cipher, \citet{adominicai} noticed that the
permutation layer can be decompose into, first, a transformation that
can be efficiently implemented within a register and, second, a
permutation that commutes with the other layers (confusion and key
mixing). Similarly, the diffusion layer of Skinny can be decomposed as
follows
%
\begin{prop}
  \coqfrom{Skinny/FixsliceProof}{split_phi_sigma}
\end{prop}
%
%% ** phi: costly permutation
%% *** -> not applied
%
where \coqe!phi_! is a permutation with inverse \coqe!inv_phi_! that
captures all the costly bit-twiddling operations
%
\coqfrom{Skinny/Base}{phi}
\coqfrom{Skinny/Base}{inv_phi}
%
%% ** sigma: efficient
%% *** <- xor: machine word xor
%% *** <- Cols.ror: machine word rotation
%% **** ARM: barrel shifter: xor + shift
%
while, in turn, \coqe!sigma_! admits an efficient implementation on
machine words
%
\coqfrom{Skinny/Base}{sigma}

@@

One solution\footnote{Needless to say, we actually started off with
the answers, kindly provided by~\citet{adomi}, and crawled our way
back to see if an alternative approach would be feasible.}, which
consists in using Coq as a term rewriting engine, is to postulate the
existence of such an explicit form, let us call it
%
\coqe!mixcolumns_mod_explicit!.
%
We then claim that this definition ought to be equivalent to its
specification, \coqe!mixcolumns_mod! for $n \in \{0, 1, 2, 3\}$.
Making sure that the operations of Boolean algebra, Naperian and
foldable functors are opaque, $\beta$ reduction yields a term in our
target language. However, it suffers from significant redundancy: by
working over the term algebra, we have not quotiented out the
equational theory. To do so, we orient the equational theory in a
rewrite database and apply the \coqe!autosubst! tactics to obtain
simplified forms. Similarly, we retrieve convenient \coqe!let! forms
thanks to the \coqe!set (ident := term)! tactics. We are then left
with reading off the actual definition of
\coqe!mixcolumns_mod_explicit!.

@@

We have thus completed our journey toward a rationalized treatment of
bitslicing and fixslicing in their quintessential form, taking the
Skinny cipher as our running example.
%
%% * Other applicative: Gift
%% ** Specification: only thinking: Rows.T Cols.T Slice4.T bool
%% *** rest is rote functionalisation
%% ** Bitslicing: extract Slice4.T out and Double
%% ** Fixslicing: decompose permutation into in-register followed by 90 degree rotation
%% *** cancel out after 4
%% *** -> simpler
%
We have similarly treated the Gift cipher, whose design is in fact at
the origin of the fixslicing technique~\citep{adomi}. Gift is
specified in terms of the composition
%
\coqe!Rows.T (Cols.T (Slice4.T bool))!
%
where \coqe!Slice4.T! denotes a depth of 4 slices. In bitsliced form,
we once again extract out the \coqe!Slice4.T! functor and
\coqe!Double.T! the amount of data processed per run. Fixslicing is
conceptually easier to understand as the diffusion layer is defined as
the combination of an in-register transformation followed by a
90-degree rotation of the matrix of rows and columns. It simply
cancels out after 4 steps.

@@

\bibliographystyle{alpha-fr}
\bibliography{paper}

>>>>

Anyone who has ever read a cryptography paper knows this for a fact:
cryptographers are at the top of the academic thought chain. In the
archetypal symmetric cryptography paper, there is a ``Section 2''
scene in which cryptographers are bravely waving their most potent
finite field. In a latter ``Section 4'' scene, the very same
cryptographers are seen extolling the virtues of some nifty
\textsc{Arm} Cortex assembly instruction, which takes center stage in
their hand-crafted, manually optimized implementation. No wonder we
jealously stash them away in ``zone à régime restrictive''! The rumor
goes that, behind these padlocked doors, they have developed a set of
domain-specific techniques to deliver very high-throughput software
implementation of symmetric ciphers. This has been slowly gnawing at
compiler writers, for whom job
security~\citep{bernstein:death-compiler,venkatesh:compiler-security}
has become a major concern. Luckily for us, every now and then,
someone inadvertently let the genies out of their Klein bottle and we
get to read about the subtle art of
``bitslicing''~\citep{biham:DES} and
``fixslicing''~\citep{adominicai:fixslicing-gift,
  adomnicai:fixslicing-AES-like}.

@@

In the long run, our ambition is to offer cryptographers the
conceptual apparatus to better illuminate their ``Section 5'' scene.
Each and every symmetric cryptosystem should not have to exhibit a
bitsliced or fixsliced implementation: by just \emph{specifying} the
cryptosystem in the right framework (abstract yet suitably
operational), one should get the guarantee that the design can be
bitsliced and fixsliced. In fact, one would boldly claim that a
compiler could automatically apply these transformations and produce
optimized code, starting from an high-level description of the cipher.
This is already the case for bitslicing~\citep{mercadier:PhD}, adding support
for fixslicing is next on the list.

@@

To avoid side-channel attack (for example, based on
timing~\citep{bernstein:timing-attack-AES}), this pipeline can be more
precisely understood as a combinational circuit. We are forbidden from
performing control-flow operations on secret data, leaving only
statically-bounded loops. We are also forbidden from performing memory
access based on the data flow, leaving only register spilling.

@@

Following the NSA involvement in the design of the DES diffusion
layer~\citep{johnson:cold-war-crypto}, cryptographers have been
advocating for ``nothing-up-my-sleeve'' designs (which does not make
magic tricks impossible, just
harder~\citep{bernstein:something-up-my-sleeves}). For example, the
diffusion layer of the AES standard is based on the usual
multiplicative inverse over the finite field induced by an
innocent-looking polynomial. To support such design, the state of
modern ciphers is thus equipped with some matricial structure
(oftentimes, manipulated as a 2D or 3D array of bits): the cipher is
not specified over a flat sequence of individual bits but over a
n-dimensional array of bits.

@@

Moreover, ciphers are now, by and large, designed to be run
efficiently on software. The era of hardware-based cryptography is
long gone, even for embedded
platforms~\citep{buchanan:nist-lightweight-crypto}. To achieve
high-throughput, ciphers are typically designed so as to be admit a
bitsliced implementation. Bitslicing enables a form of parallel
execution \emph{within} machine words. We thus gain access to a form
of CPU-level ``scale out'' to larger machine words, in particular
through single-instruction multiple-data (SIMD) instructions. Doubling
the size of machine words yields nearly twice the throughput, without
any effort. Evidences tend to suggest that cryptographers obtain
bitsliced designs through sheer intellectual might. It is in fact the
sole purpose of the ``Section 5'' scene to exhibit a bitsliced witness
in all the gory details.

@@

We believe that there is also a broader take-away for an audience of
stubbornly French functional programmers. The present work surfs on
the wake of the Squiggol school of
programming~\citep{bird:squiggol,roconnor:functor-oriented}. Because our
government-approved programming language does not yet support ad-hoc
polymorphism~\citep{white:modular-implicit}, the French community may be
missing out on some interesting programming patterns. The present
article can thus be read as a case-study in ``functor-oriented
programming'', where the word ``functor'' is taken in his categorical
meaning (recalled in Section~\ref{sec:spec} and unrelated to the one
mentioned in the ML module system). In effect, we forbid ourselves
from programming over a concrete datastructure: instead, we rely on
(categorical) functors, peppered with some more structure, so as to 1.
unlock algebraic manipulation during compilation and 2. exploit
parametricity for reasoning.

@@

The present work is but the beginning of our research program. We
shall focus \emph{exclusively} on the semantics aspects, only briefly
touching upon implementation aspects (which remain the raison d'être
of the project!). Our experience designing the Usuba~\citep{mercadier:PhD}
programmign language and implementing its compiler gives us some
confidence that a syntax can indeed be tailored to suit this
semantics.

We nonetheless could not resist the temptation of developing the
semantics in the Coq theorem prover, as executable programs. After
all, we are interested in combinational circuits: if there has ever
been a time where we do not have to worry about termination, this is
it! So we have implemented cryptographical primitive in pure Gallina.
This is nothing to worry about: after all, more ambitious colleagues
went from there down to deeply-embedded assembly
code~\citep{chlipala:coq-crypto}. Besides the ability to prove the
correctness and, even more usefully, test our code, this also enables
us to easily prototype a fixslicing compiler, using Coq's
\texttt{autorewrite} tactics to emulate a transformation-based simplification
engine~\citep{peyton-jones:haskell-optim} (part of an imaginary optimizing
compiler).

@@

This article grew out of the Bachelor's research project of the first
author and the Master thesis of the second author. The former worked
on the specification, bitslicing and fixslicing of the Skinny
cipher~\citep{beierle:skinny, adomnicai:fixslicing-AES-like}. The
latter worked on the specification, bitslicing and fixslicing of the
Gift cipher~\citep{subhadeep:gift, adominicai:fixslicing-gift}. Both
ciphers are provided in the aaccompanying source
code~\footnote{\url{https://github.com/pedagand/bitfix}}. For clarity,
we shall focus exclusively on Skinny as our running example here.

@@

Instead, we require \coqe!Rows.T! to offer a \coqe!map! operation: it
constructively witnesses the fact that a \coqe!Rows.T A! is
(functionally) related to a \coqe!Rows.T B!, as long as we can
(functionally) relate \coqe!A! and \coqe!B!. Put otherwise, we ask
that \coqe!Rows.T!, and similarly \coqe!Cols.T! and \coqe!Slice8.T!,
come equipped with their free theorems~\citep{reynolds:parametricity,
  wadler:free-theorem}.

@@

Much like $\lambda$ in functional programming, the exclusive-or
features prominently in the cryptographic liturgy. In particular, it
is instrumental to mix the derived keys during the ciphertext
computation. For genericity, we model this layer as an operation
defined over any Boolean algebra~\citep{givant:boolean-algebras}:
%
\coqfrom{Skinny/Base}{addroundkey}

@@

The type \coqe!bool! is an obvious instance of a Boolean algebra.
However, any commutative applicative~\citep{mcbride:applicative}
functor \coqe!F! applied to a Boolean algebra also yields a Boolean
algebra (dispatching the Boolean operations pointwise to the
underlying elements). For the sake of completeness, we recall that an
applicative functor offers the following operations:
%
\coqfrom{Algebra}{Applicative}

@@

The role of the diffusion layer is to divert individual bits across the
whole structure. Inevitably, this calls for a notion of indexing.
Naperian functors~\citep{hancock:napierian} identify a type \coqe!Ix! of
indices as the logarithm of an exponential type:
%
\coqfrom{Algebra}{Naperian}
%
through the fact that \coqe!lookup! and \coqe!init! form a bijection.
This is also known as a representable functor in mathematical
circles~\citep{maclane:working-categorist}

@@

The first step of diffusion process consists in performing a right
rotation by a specific amount over each individual column. However, we
do not have enough structure to identify a ``right'' or ``left''
direction over our containers. The standard tool of the trade to
achieve this effect is the notion of foldable
functor~\citep{yorgey:typeclassopedia}:
%
\coqfrom{Algebra}{Foldable}

@@

\begin{figure}[htp]
  \centering
  \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \includegraphics[draft=false,height=2.8cm]{shiftrows}
    \caption{Shiftrows (from~\citet{beierle:skinny})}
    \label{fig:shiftrows}
  \end{subfigure}
  \begin{subfigure}[t]{0.45\textwidth}
    \centering
    \includegraphics[draft=false,height=2.8cm]{mixcolumns}
    \caption{Mixcolumns (from~\citet{beierle:skinny})}
    \label{fig:mixcolumns}
  \end{subfigure}
\end{figure}

@@

\begin{figure}
  \centering
  \includegraphics[draft=false,height=5cm]{s8}
  \caption{Substitution box S8 (from~\citet{beierle:skinny})}
  \label{fig:s8}
\end{figure}

@@

Given two Naperian functors \coqe!F! and \coqe!G!, there exists a
generic notion of reindexing~\citep{gibbons:apl}:
%
\coqfrom{Algebra}{reindex}

@@

Studying the Gift cipher, \citet{adominicai:fixslicing-gift} noticed that the
permutation layer can be decompose into, first, a transformation that
can be efficiently implemented within a register and, second, a
permutation that commutes with the other layers (confusion and key
mixing). Similarly, the diffusion layer of Skinny can be decomposed as
follows
%
\begin{prop}
  \coqfrom{Skinny/FixsliceProof}{split_phi_sigma}
\end{prop}
%
%% ** phi: costly permutation
%% *** -> not applied
%
where \coqe!phi_! is a permutation with inverse \coqe!inv_phi_! that
captures all the costly bit-twiddling operations
%
\coqfrom{Skinny/Base}{phi}
\coqfrom{Skinny/Base}{inv_phi}
%
%% ** sigma: efficient
%% *** <- xor: machine word xor
%% *** <- Cols.ror: machine word rotation
%% **** ARM: barrel shifter: xor + shift
%
while, in turn, \coqe!sigma_! admits an efficient implementation on
machine words
%
\coqfrom{Skinny/Base}{sigma}

@@

One solution\footnote{Needless to say, we actually started off with
the answers, kindly provided by~\citet{adominicai:fixslicing-gift}, and crawled our way
back to see if an alternative approach would be feasible.}, which
consists in using Coq as a term rewriting engine, is to postulate the
existence of such an explicit form, let us call it
%
\coqe!mixcolumns_mod_explicit!.
%
We then claim that this definition ought to be equivalent to its
specification, \coqe!mixcolumns_mod! for $n \in \{0, 1, 2, 3\}$.
Making sure that the operations of Boolean algebra, Naperian and
foldable functors are opaque, $\beta$ reduction yields a term in our
target language. However, it suffers from significant redundancy: by
working over the term algebra, we have not quotiented out the
equational theory. To do so, we orient the equational theory in a
rewrite database and apply the \coqe!autosubst! tactics to obtain
simplified forms. Similarly, we retrieve convenient \coqe!let! forms
thanks to the \coqe!set (ident := term)! tactics. We are then left
with reading off the actual definition of
\coqe!mixcolumns_mod_explicit!.

@@

We have thus completed our journey toward a rationalized treatment of
bitslicing and fixslicing in their quintessential form, taking the
Skinny cipher as our running example.
%
%% * Other applicative: Gift
%% ** Specification: only thinking: Rows.T Cols.T Slice4.T bool
%% *** rest is rote functionalisation
%% ** Bitslicing: extract Slice4.T out and Double
%% ** Fixslicing: decompose permutation into in-register followed by 90 degree rotation
%% *** cancel out after 4
%% *** -> simpler
%
We have similarly treated the Gift cipher, whose design is in fact at
the origin of the fixslicing
technique~\citep{adominicai:fixslicing-gift}. Gift is specified in
terms of the composition
%
\coqe!Rows.T (Cols.T (Slice4.T bool))!
%
where \coqe!Slice4.T! denotes a depth of 4 slices. In bitsliced form,
we once again extract out the \coqe!Slice4.T! functor and
\coqe!Double.T! the amount of data processed per run. Fixslicing is
conceptually easier to understand as the diffusion layer is defined as
the combination of an in-register transformation followed by a
90-degree rotation of the matrix of rows and columns. It simply
cancels out after 4 steps.

@@

\bibliographystyle{unsrtnat}
\bibliography{paper}
